---
title: "ROMA BL samples"
author: "marie"
date: "21 mai 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

To install Roma 
https://github.com/sysbio-curie/rRoma

https://rdrr.io/github/Albluca/rRoma/man/rRoma.R.html

```{r rRoma}
library(readxl)
library(rRoma)
library(DESeq2)
library(ggplot2)
library(corrplot)
library(tidyverse)
library(viridis)   
library(ggthemes) 
```



```{r load, echo=FALSE}

metadataTot1=read.table(file="metadataTot_FINAL.txt")
CountTot=read.table(file="CountTot.txt")

ddsToothWhole <- DESeqDataSetFromMatrix(countData = round(CountTot[,-c(grep("kidney",metadataTot1$machoire))],0),
                              colData = metadataTot1[-c(grep("kidney",metadataTot1$machoire)),],
                              design = ~ 1)

colData(ddsToothWhole)$stade=as.numeric(as.character(colData(ddsToothWhole)$stade))
colData(ddsToothWhole)$timerel[colData(ddsToothWhole)$espece=="mus"]=10*(as.numeric(colData(ddsToothWhole)$stade[colData(ddsToothWhole)$espece=="mus"])-14.5)/(18-14.5)
colData(ddsToothWhole)$timerel[colData(ddsToothWhole)$espece=="ham"]=10*(as.numeric(colData(ddsToothWhole)$stade[colData(ddsToothWhole)$espece=="ham"])-12)/(14.5-12)
colData(ddsToothWhole)$machoire=as.factor(colData(ddsToothWhole)$machoire)                              
                              
ddsToothWhole <- DESeq(ddsToothWhole)
vstT=data.frame(assay(vst(ddsToothWhole)))
bm=data.frame(counts(ddsToothWhole,norm=T))

 ```




```{r pathways, echo=FALSE}

bmp=read.csv("BMP-avril2019.csv")
bmp$dir="not target"
bmp$dir[grep("negative",bmp$X)]="negative"
bmp$dir[grep("positive",bmp$X)]="positive"
## clean
bmp=bmp[!(bmp$Target=="Wnt3a"&bmp$X=="BMP EPI EPI positive target"),]

bmpE10=unique(bmp[bmp$Reference=="This study"&bmp$Stage=="E10",])
bmpE13E14=unique(bmp[bmp$Reference=="This study"&(bmp$Stage=="E13"|bmp$Stage=="E14"),])
bmpE13E14=unique(bmpE13E14[,c("Target","dir")])


MesModulesBmpE10 = list()
MesModulesBmpE10[["BmpE10"]]=list(Name="BmpE10",
                         Desc="BmpE10",
                         Genes=bmpE10$Target,
                         Weigths=ifelse(bmpE10$dir=="positive",1,-1))

MesModulesBmpE13E14 = list()
MesModulesBmpE13E14[["BmpE13E14"]]=list(Name="BmpE13E14",
                         Desc="BmpE13E14",
                         Genes=bmpE13E14$Target,
                         Weigths=ifelse(bmpE13E14$dir=="positive",1,-1))




wnt=read.csv("WNT-avril2019.csv")
wnt$dir="not target"
wnt$dir[grep("negative",wnt$X)]="negative"
wnt$dir[grep("positive",wnt$X)]="positive"
wnt=wnt[!(wnt$Target=="Wnt10a"&wnt$X=="EPI EPI negative target"),]

wntE10=wnt[wnt$Reference=="This study"&wnt$Stage=="E10",]
wntE13E14=wnt[wnt$Reference=="This study"&(wnt$Stage=="E13"|wnt$Stage=="E14"),]
wntE13E14=unique(wntE13E14[,c("Target","dir")])

wntAll=wnt[(wnt$Stage=="E13"|wnt$Stage=="E14"),]
wntAll=unique(wntAll)

MesModulesWntE10 = list()
MesModulesWntE10[["WntE10"]]=list(Name="WntE10",
                         Desc="WntE10",
                         Genes=wntE10$Target,
                         Weigths=ifelse(wntE10$dir=="positive",1,-1))

MesModulesWntE13E14 = list()
MesModulesWntE13E14[["WntE13E14"]]=list(Name="WntE13E14",
                         Desc="WntE13E14",
                         Genes=wntE13E14$Target,
                         Weigths=ifelse(wntE13E14$dir=="positive",1,-1))


# liste SHH dans:
# https://www.gsea-msigdb.org/gsea/msigdb/cards/PID_HEDGEHOG_GLI_PATHWAY
listShh=read.table("genesetSHH.txt",h=F)
c=data.frame(cbind(row=row.names(ddsToothWhole),rowup=toupper(row.names(ddsToothWhole))))
listShht=c[c$rowup%in%listShh$V1,]
# On met activation pour tous les gènes
MesModulesShh = list()
MesModulesShh[["Shh"]]=list(Name="Shh",
                          Desc="Shh",
                          Genes=listShht$row,
                          Weigths=rep(1,length(listShht$row)))




fcorSp4=function(species="mus",set="bmpE10",MesModulesBW=MesModulesBmpE10, optPC="CorrelateKnownWeightsByGene",CorM="spearman",nb=nb)
{
vstTmd=vstT[,colData(ddsToothWhole)$espece==species]
cmd=colData(ddsToothWhole)[colData(ddsToothWhole)$espece==species,]
cmd$rep[grep("4.",row.names(cmd))]="rep4"
cmd$rep[grep("3.",row.names(cmd))]="rep3"
cmd$type="whole"
cmd$type[grep("epi",cmd$machoire)]="epi"
cmd$type[grep("mes",cmd$machoire)]="mes"
cmd$type[grep("_B",cmd$machoire)]="B"
cmd$type[grep("_L",cmd$machoire)]="L"
cmd$machoire[grep("md",cmd$machoire)]="md"
cmd$machoire[grep("mx",cmd$machoire)]="mx"
vstTmd=vstTmd[,cmd$type=="B"|cmd$type=="L"]
cmd=cmd[cmd$type=="B"|cmd$type=="L",]

names(vstTmd)=paste0(cmd$machoire,"_",cmd$stade,"_",cmd$rep,"_",cmd$type)
  Data_mus_md <- rRoma.R(ExpressionMatrix = vstTmd, ModuleList = MesModulesBW, FixedCenter = TRUE,
                   UseParallel = TRUE, nCores = 3, ClusType = "FORK", MaxGenes = 100,MinGenes = 4,
                   PCSignMode=optPC, CorMethod= CorM,UseWeigths = TRUE,OutGeneNumber=10,SuppressWarning = TRUE,MoreInfo=FALSE)

Data_mus_md_score=data.frame(sapply(1,function(x){as.numeric(Data_mus_md$ModuleSummary[[x]]$SampleScore)}))
names(Data_mus_md_score)=set
row.names(Data_mus_md_score)=names(vstTmd)

weight=data.frame(weight=Data_mus_md$ModuleSummary[[1]]$GeneWeight*Data_mus_md$ModuleSummary[[1]]$CorrectSign1)
weight2=data.frame(weight=MesModulesBW[[1]]$Weigths)
weight2$name=MesModulesBW[[1]]$Genes
weight2=merge(weight,weight2,by.x=0,by.y="name")
names(weight2)=c("name","weight","activ")

weight2=weight2[order(weight2$weight),]
neg=sum(weight2[1:nb,"activ"]>0)
pos=sum(weight2[c((nrow(weight2)-nb):nrow(weight2)),"activ"]>0)

if(neg > pos){
  Data_mus_md_score[,1]=-Data_mus_md_score[,1];
  cat("sign changed \n")
}
cat(paste0("negative side: ",species," ", neg," activator. positive side: ",pos,"\n"))
return(Data_mus_md_score)
}




fgatherSp4=function(MesModulesBWx=MesModulesBmpE10,setx="bmpE10",optPC="CorrelateKnownWeightsByGene",CorMethod="pearson",nb=10)
{
Data_mus_score_bmpE10=fcorSp4(species="mus",set=setx,MesModulesBW=MesModulesBWx,optPC=optPC,CorM=CorMethod, nb=nb)

library(tidyr)
Data_mus_score_bmpE10$id=row.names(Data_mus_score_bmpE10)

Data_mus_score_long=gather(Data_mus_score_bmpE10,key = "pathway", value = "value", -id)
Data_mus_score_long$species="mus"

Data_mus_score_long$tooth="md"
Data_mus_score_long$tooth[grep("mx",Data_mus_score_long$id)]="mx"

Data_mus_score_long$type="whole"
Data_mus_score_long$type[grep("epi",Data_mus_score_long$id)]="epi"
Data_mus_score_long$type[grep("mes",Data_mus_score_long$id)]="mes"
Data_mus_score_long$type[grep("B",Data_mus_score_long$id)]="B"
Data_mus_score_long$type[grep("L",Data_mus_score_long$id)]="L"

Data_mus_score_long$stage=as.numeric(sapply(strsplit(Data_mus_score_long$id,"_"),function(x){x[[2]]}))
Data_mus_score_long$rep=sapply(strsplit(Data_mus_score_long$id,"_"),function(x){x[[3]]})

dataROMA=data.frame(Data_mus_score_long)

g=ggplot(dataROMA, aes(as.factor(stage), type, fill= value)) + 
  geom_tile()+ facet_grid(tooth ~ species, scales="free_x")+
  scale_fill_gradient(low = "white", high = "red")
ggsave(g,file=paste0("dataROMA_orient2_","_",setx,"_",optPC,"_",CorMethod,".png"))

return(dataROMA)
}


setx="bmpE10"
f1d=fgatherSp4(MesModulesBWx=MesModulesBmpE10,setx="bmpE10",optPC="none",CorMethod="pearson",nb=5)
g=ggplot(f1d, aes(tooth, value, color= type)) + 
  geom_point()
ggsave("BmpE10ROMAd.png",g)


setx="bmpE13E14"
f2d=fgatherSp4(MesModulesBWx=MesModulesBmpE13E14,setx="bmpE13E14",optPC="none",CorMethod="pearson",nb=5)
g=ggplot(f2d, aes(tooth, value, color= type)) + 
  geom_point() 
ggsave("BmpE13E14ROMAd.png",g)

setx="wntE10"
f4d=fgatherSp4(MesModulesBWx=MesModulesWntE10,setx="wntE10",optPC="none",CorMethod="pearson",nb=5)
g=ggplot(f4d, aes(tooth, value, color= type)) + 
  geom_point()
ggsave("WntE10ROMAd.png",g)

setx="wntE13E14"
f5d=fgatherSp4(MesModulesBWx=MesModulesWntE13E14,setx="wntE13E14",optPC="none",CorMethod="pearson",nb=5)
g=ggplot(f5d, aes(tooth, value, color= type)) + 
  geom_point()
ggsave("WntE13E14ROMAd.png",g)


setx="Shh"
f6d=fgatherSp4(MesModulesBWx=MesModulesShh,setx="shh",optPC="none",CorMethod="pearson",nb=5)
g=ggplot(f6d, aes(tooth, value, color= type)) + 
  geom_point()
f6d$value=-f6d$value
g=ggplot(f6d, aes(tooth, value, color= type)) +   geom_point()
ggsave("ShhROMAd.png",g)



tabBLRoma=data.frame(rbind(f1d,f2d,f4d,f5d,f6d))



tabBLRoma$pathComp=tabBLRoma$pathway
tabBLRoma$pathComp=gsub("E10"," (Mes)",tabBLRoma$pathComp)
tabBLRoma$pathComp=gsub("E13E14"," (Epi)",tabBLRoma$pathComp)

pathComp <- c("MesBmp","EpiBmp","MesWnt","EpiWnt","Shh")
names(pathComp) <- c("bmp_Mes","bmp_Epi","wnt_Mes","wnt_Epi","shh")


gg=ggplot(tabBLRoma, aes(tooth, value, color= type)) + geom_point() +  ylab("Activation level") + 
scale_x_discrete(name ="Tooth",labels=c("md"="lower","mx"="upper"))+
  scale_color_manual(values=c("#E69F00", "#56B4E9"),labels=c("B"="buccal","L"="lingual"))+
  guides(color=guide_legend(title="Tooth side")) + 
facet_grid( ~ pathComp, scales="free_x", labeller = labeller(pathComp=pathComp)) 
ggsave("tabBLRoma.png",gg)
ggsave("tabBLRoma.pdf",gg)

tabBLRomaM <- tabBLRoma %>% group_by(pathComp,tooth,type) %>% 
      summarise(mean_activ = mean(value))
g=ggplot(tabBLRomaM, aes(tooth, mean_activ, color= type)) + geom_point() +  ylab("Activation level") + 
scale_x_discrete(name ="Tooth",labels=c("md"="lower","mx"="upper"))+
  scale_color_manual(values=c("#E69F00", "#56B4E9"),labels=c("B"="buccal","L"="lingual"))+
  guides(color=guide_legend(title="Tooth side")) + 
facet_grid( ~ pathComp, scales="free_x", labeller = labeller(pathComp=pathComp)) 
ggsave("tabBLRoma_Average.png",g)
ggsave("tabBLRoma_Average.pdf",g)


g=ggplot(tabBLRomaM, aes(y=tooth, fill=mean_activ, x= type)) + geom_tile(color="white", size=0.1)+
scale_y_discrete(name ="Tooth",labels=c("md"="lower","mx"="upper"))+
scale_x_discrete(name ="Tooth Part",labels=c("B"="buccal","L"="lingual"))+
  scale_fill_viridis(name="Activation",direction=-1)+
  coord_equal()+
  facet_grid( ~ pathComp, labeller = labeller(pathComp=pathComp)) +
  theme(axis.ticks=element_blank())+
  theme(legend.text=element_text(size=6))+ ggpubr::rotate_x_text()
ggsave("tabBLRoma_tiles.png",g)
ggsave("tabBLRoma_tiles.pdf",g)

g2=ggplot(tabBLRomaM, aes(y=tooth, fill=mean_activ, x= type)) + geom_tile(color="white", size=0.1)+
scale_y_discrete(name ="Tooth",labels=c("md"="lower","mx"="upper"))+
scale_x_discrete(name ="Tooth Part",labels=c("B"="buccal","L"="lingual"))+
  scale_fill_viridis(name="Activation",direction=-1,option="G")+
  coord_equal()+
  facet_grid( ~ pathComp, labeller = labeller(pathComp=pathComp)) +
  theme(axis.ticks=element_blank())+
  theme(legend.text=element_text(size=6))+ ggpubr::rotate_x_text()
ggsave("tabBLRoma_tiles2.png",g2)
ggsave("tabBLRoma_tiles2.pdf",g2)



g3 = ggarrange(g2, gg , 
          labels = c("A", "B"),
          ncol = 1, nrow = 2)
ggsave("tabBLRoma_tiles3.png",g3)
